Open Policing Project Tutorial

### 2019-03-04 nicar_tutorial.Rmd
progress

---
title: "Open Policing Project Tutorial"
output: html_document
---
NOTES: We don't really have enough data in Jan 2014 to trust. Not sure whether
to mention this or if it just complicates things too much, since eventually
we just filter to 2017. But it is a little bit relavent for the by-year plot...
we'd probably expect to see about 15,000 or so more stops (which, spread across
races is a drop in the bucket) -- honestly wouldn't change the plot much; it's
just a matter of thoroughness/integrity.

Things that i use but don't explain:

* assignment operator ( <- )
* pipe ( %>% )
* filter
* mutate
* group_by/summarize
* select
* left_join
* super basic ggplot (i.e., aes(), geom_point(), geom_line())

@@

## Load the data
# Replace the path below with the path to where your data lives
data_path <- "~/opp/data/states/pa/philadelphia/clean/philadelphia.rds"
stops <- read_rds(data_path)$data

@@

@@

Now, you may have noticed that there was a column called `type`. Let's take a
closer look to see what that means.

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
stops %>% 
  head(10) %>% # this allows us to look at just the first 10 rows
  select(type)
```

@@

It'd be nice to see if the 1.1 million stops were evenly distributed across 
years, or if stop counts have changed. To find stop counts per year, we need to
define a notion of `year` (our data only has `date`). Luckily, with the 
`lubridate` package, it's as easy as `year(date)`!

@@

At first glance, we see there are for more stops of black drivers than any other
race. About two-thirds of stops in our dataset were of black drivers! This 
stat on is own, though, doesn't actually say much. We'll return to this more
rigorously later on.

@@

*Fun fact:* This same trend does _not_ hold true for pedestrian stops. In fact,
if we hadn't filtered to only vehicular stops, we wouldn't have noticed this,
because this same plot over the combined pedestrian + vehicular data looks
unremarkable. The takeaway is that looking at trends by sub-categories can often
be very helpful. (e.g., In Nashville, looking at the different listed stop 
reasons uncovers the extent of the disparities.)

To give us time to dive into a variety of analysis methods, we'll let you 
investigate Philly's annual traffic stop patterns on your own time. But because
it seems like the disparities could be changing over time, we'll filter to
only 2017 and focus on that data for the rest of our analysis.

@@

(*Fun fact:* The threshold test confirms the results we see in Philadelphia.
Officers are indeed applying lower thresholds when deciding to search black and 
Hispanic drivers than when deciding to search white drivers.)

@@

TODO(amyshoe): Discuss setup of vod test, walk through first code chunk.

@@

# Compute sunset time for each date in our dataset
sunset_times <- 
  stops %>%
  mutate(
    lat = center_lat,
    lon = center_lng
  ) %>% 
  select(date, lat, lon) %>%
  distinct() %>%
  mutate(
    sunset = format(
      getSunlightTimes(data = ., keep = c("sunset"), tz = tz)$sunset,
      "%H:%M:%S"
    ), 
    sunset_minute = time_to_minute(sunset)
  ) %>% 
  select(date, sunset, sunset_minute)

vod_stops <- 
  stops %>% 
  left_join(
    sunset_times,
    by = "date"
  ) %>% 
  mutate(
    minute = time_to_minute(time),
    minutes_after_sunset = minute - sunset_minute,
    is_dark = minute > sunset_minute,
    min_sunset_minute = min(sunset_minute),
    max_sunset_minute = max(sunset_minute),
    is_black = subject_race == "black"
  ) %>% 
  filter(
    # NOTE: filter to get only the intertwilight period
    minute >= min_sunset_minute,
    minute <= max_sunset_minute,
    subject_race %in% c("black", "white")
  )
```

TODO(amyshoe): Talk about controlling for time, and the plot as intuition.
Walk through second code chunk.

```{r}
to_quarter_hour <- function(date, time) {
  format(
    round_date(ymd_hms(str_c(date, time, sep = " ")), "15 min"),
    "%H:%M:%S"
  )
}
vod_stops %>%
  # This time range contains stops that occur in the (-1hr, 1hr) range around
  # sunset (allowing us to compare stops that occurred at the same time but
  # in darkness vs in light)
  filter(time >= hm("17:38"), time <= hm("19:22")) %>% 
    # NOTE: controlling for time every 15 minutes
    mutate(
      quarter_hour = to_quarter_hour(date, time),
      quarter_hour_minute = time_to_minute(quarter_hour),
      quarter_hour_minute_since_sunset = quarter_hour_minute - sunset_minute,
      quarter_hour_readable = str_c(
          hour(hms(quarter_hour)) - 12,
          ':',
          str_pad(minute(hms(quarter_hour)), 2, side = "right", pad = "0"),
          " PM"
      )
    ) %>%
  group_by(
    quarter_hour_readable,
    quarter_hour_minute_since_sunset
  ) %>%
  summarize(
    minority_total = sum(is_black),
    majority_total = sum(!is_black),
    proportion_minority = minority_total / (minority_total + majority_total)
  ) %>%
  ggplot(aes(
    x = quarter_hour_minute_since_sunset,
    y = proportion_minority,
    color = quarter_hour_readable
  )) +
  geom_smooth(method = "lm", se = F) +
  xlab("Minutes Since Sunset") +
  ylab("Proportion Minority") +
  coord_cartesian(xlim = c(-60, 60)) +
  theme(legend.title = element_blank())
```

TODO(amyshoe): Provide plot interpretation.
Discuss how to go from this intuition to a more rigorous model.
Walk through third code chunk.

```{r}
mod <- glm(
  is_black ~ is_dark + splines::ns(minute, df = 6),
  family = binomial,
  data = vod_stops
)

summary(mod)$coefficients["is_darkTRUE", c("Estimate", "Std. Error")]
exp(summary(mod)$coefficients["is_darkTRUE", "Estimate"])
exp(confint(mod, parm = "is_darkTRUE"))
```

TODO(amyshoe): Interpret model results
Talk about robustness checks (adjust for sub-geogrphy, vary spline degree etc)

```{r}
mod <- glm(
  is_black ~ is_dark + splines::ns(minute, df = 6) + as.factor(district),
  family = binomial,
  data = vod_stops
)

summary(mod)$coefficients["is_darkTRUE", c("Estimate", "Std. Error")]
# confint(mod, parm = "is_darkTRUE")
```

>>>>

---
title: "Open Policing Project Tutorial"
output: html_document
---

@@

## Load the data
# Replace the path below with the path to where your data lives
data_path <- "~/opp/data/states/pa/philadelphia/clean/philadelphia.rds"
stops <- read_rds(data_path)#$data

@@

To do this we'll want to

1. use the `filter()` function to specify the years we want, and 
2. use the assignment operator `<-` (it's like = in `R`) to overwrite `stops`.

You might have noticed in the colnames, that we have no `year` column (our data 
only has `date`). Luckily, with the `lubridate` package, getting the year from
the date is as easy as `year(date)`!

@@

```{r}
# NOTE that this is equivalent to `nrow(stops)`
stops %>% nrow()
```

**tidyverse tip:** The "pipe" operator, `%>%`, just places the previous item
into the first argument of the function. So, `x %>% f()` is simply `f(x)`. While
in a one-function call the pipe might feel silly and unnecessary, it's going to
become _really_ helpful once we start wanting to do multiple transformations to
our data.

Now, you may have noticed before that there was a column called `type`. Let's 
take a closer look to see what that means.

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
stops %>% 
  select(type) %>% 
  head(10) # this allows us to look at just the first 10 rows
```

@@

It'd be nice to see if the 1.1 million stops were evenly distributed across 
years, or if stop counts have changed. To find stop counts per year, we need to
define a notion of `year` (recall that our data only has `date`).

@@

At first glance, we see there are far more stops of black drivers than any other
race. About two-thirds of stops in our dataset were of black drivers! This 
stat on is own, though, doesn't actually say much. We'll return to this more
rigorously later on.

@@

**Fun fact:** This same trend does _not_ hold true for pedestrian stops. In fact,
if we hadn't filtered to only vehicular stops, we wouldn't have noticed this,
because this same plot over the combined pedestrian + vehicular data looks
unremarkable. The takeaway is that looking at trends by sub-categories can often
be very helpful. (e.g., In Nashville, looking at the different listed stop 
reasons uncovers the extent of the disparities.)

To give us time to dive into a variety of analysis methods, we'll let you 
investigate Philly's annual traffic stop patterns on your own time. But because
it seems like the disparities could be changing over time, we'll filter to
only 2017 and focus on that data for the rest of our analysis (and overwrite
`stops`).

@@

(**Fun fact:** The threshold test confirms the results we see in Philadelphia.
Officers are indeed applying lower thresholds when deciding to search black and 
Hispanic drivers than when deciding to search white drivers.)

@@

The outcome test is a good way to get around the benchmarking problem at the
search decision, but when it comes to assessing bias at the stop decision,
it's a little trickier. The outcome test usually doesn't work, since rarely
is there a notion of what a "successful" stop is. (There are, of course, some
exceptions to this, like when stop reason is listed as suspected Criminal 
Possession of a Weapon (CPW) -- which is the case in NYC stop-question-frisk
data.)

An alternative approach to assessing bias in stop decisions was proposed
by Grogger and Ridgeway in 2006. This approach, known as
the Veil of Darkness test, relies on the hypothesis that officers who are
engaged in racial profiling are less likely to be able to identify a driver's
race after dark than during daylight. Under this hypothesis, if stops made after dark had smaller proportion of black drivers stopped than stops made during daylight, this would be evidence of the presence of racial profiling.

The naive approach is just to compare whether daytime stops or nighttime stops 
have higher proportion of black drivers. This, however, is problematic. More black
drivers being stopped at night could be the case for a multitude of reasons: 
different deployment and enforcement patterns, different driving patterns,
etc. All of these things correlate with clock time and would be confounding our
results. So let's find a way to adjust for clock time.

To adjust for clock time, we want to compare times that were light at some
point during the year (i.e., occurred before dusk) but were dark at other points
during there (i.e., occurred after dusk). This is called the "inter-twilight
period": the range from the earliest time dusk occurs in the year to the latest 
time dusk occurs in the year.

Let's calculate the sunset and dusk times for Philly in 2017.

(Note that we distinguish here between _sunset_ and _dusk_. Sunset is the point 
in time where the sun dips below the horizon. Dusk, also known as the end of 
civil twilight, is the time when the sun is six degrees below the horizon, and 
is it is widely considered to be dark. We'll explain why this is relevant in a
moment.)

@@

# Compute sunset time for each date in our dataset
sunset_times <- 
  stops %>%
  mutate(
    lat = center_lat,
    lon = center_lng
  ) %>% 
  select(date, lat, lon) %>%
  distinct() %>%
  getSunlightTimes(
    data = ., 
    keep = c("sunset", "dusk"), 
    tz = tz
  ) %>% 
  mutate_at(vars("sunset", "dusk"), ~format(., "%H:%M:%S")) %>% 
  mutate(
    sunset_minute = time_to_minute(sunset),
    dusk_minute = time_to_minute(dusk),
    date = ymd(str_sub(date, 1, 10))
  ) %>% 
  select(date, sunset, dusk, ends_with("minute"))
```

Great. Now, using `sunset_times`, what is Philly's inter-twilight period?

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
sunset_times %>% 
  filter(dusk == min(dusk) | dusk == max(dusk))
```

So the earliest sunset in 2017 was at around 4:30pm, in December (and it was
fully dark about 30 minutes later, at dusk, around 5pm). The latest sunset time
in 2017 was at around 8:30pm, in late June (and it was fully dark about 30 
minutes later, at dusk, around 9pm).

For a time in this range, like 6:30pm, part of the year it's still light, and part
of the year, it's still dark. So we can compare the proportion of black drivers 
stopped at 6:30pm when it's light (usually in the summer) to the proportion of 
black drivers stopped at 6:30pm when it's dark (usually in the winter). This way,
we're really seeing the effect of darkness, since driving and deployment 
patterns tend to change only with clock time. For example, the demographics
of commuters who are on the road at 6:30pm likely stays pretty steady throughout
the year, compared to demographics of commuters at 5pm versus at 9pm.

One quick note: We want to make sure to filtering out stops that occurred in the
approximately 30-minute period between sunset and dusk (end of civil twilight),
since it is ambiguous as to whether that period is light or dark, which would
muddle up our analysis.

Let's join the sunset times to our stops data and

* filter to the inter-twiligh period
* remove that ambigous pre-dusk period
* for simplicity, compare only black and white drivers

```{r}
vod_stops <- 
  stops %>% 
  left_join(
    sunset_times,
    by = "date"
  ) %>% 
  mutate(
    minute = time_to_minute(time),
    minutes_after_dark = minute - dusk_minute,
    is_dark = minute > dusk_minute,
    min_dusk_minute = min(dusk_minute),
    max_dusk_minute = max(dusk_minute),
    is_black = subject_race == "black"
  ) %>% 
  filter(
    # Filter to get only the intertwilight period
    minute >= min_dusk_minute,
    minute <= max_dusk_minute,
    # Remove ambigous period between sunset and dusk
    !(minute > sunset_minute & minute < dusk_minute),
    # Compare only whites and blacks
    subject_race %in% c("black", "white")
  )
```

How many stops are in our this new, filtered data frame?

```{r}
vod_stops %>% nrow()
```

Now, to get some intuition, let's use our 6:30pm example. If we filter the data
to stops that occurred in the narrow window from 6:30pm to 6:45pm, we can compute
what proportion of stops were of black drivers for stops when it was dark and 
compare that to the proportion for stops when it was not dark yet.

```{r}
vod_stops %>% 
  filter(time > hm("18:30"), time < hm("18:45")) %>% 
  group_by(is_dark) %>% 
  summarize(prop_black = mean(is_black))
```

~~
**NOTE FOR ERIC/JUSTIN/CHERYL/BETHANY:** I can add in another raw data intuition
section that's a basic raw data plot if you think the concept of the VOD test
is still confusing. But i want to keep things a bit briefer that way we have 
more time to field questions and have them actually type things in and think 
things through rather than just running code dan and i wrote...let me know what
you think though.
~~

Indeed we see that the proportion is smaller when 6:30-6:45pm is dark than when
it's light. According to the veil of darkness hypothesis, this is because 
officers engaged in racial profiling can't determine drivers' races as well
at night.  We know that the proportion of black drivers stopped dropped by 
about 5% when it was dark and race was harder to identify. It's hard to say what
this means in terms of the extent or pervasiveness, though. We **cannot** make
*any* claims about whether some small percent of the officers were engaged in 
stops that were 100% profiled on race, or whether all officers were engaged in
racial profiling in some small percentage of their stops, or anything of the
sort.

Furthermore, while this stat gives us some intuition that racial profiling might 
be present in Philly, we've only looked at one time. In order to make sure the
result is robust, we want to check this for _every_ time in our inter-twilight
period. To do this, we can use _logistic regression_. The basic idea here is
that we're finding what our data implied about how darkness and clock time
influence the whether a stopped driver will be black. We then extract the 
coefficient of darkness (you can think of this loosely as the measure of how
much darkness -- as opposed to clock time -- influences whether a stopped driver
will be black).

```{r}
mod1 <- glm(
  is_black ~ is_dark + splines::ns(minute, df = 6),
  family = binomial,
  data = vod_stops
)

summary(mod1)$coefficients["is_darkTRUE", c("Estimate", "Std. Error")]
#exp(summary(mod1)$coefficients["is_darkTRUE", "Estimate"])
```

The fact that the coefficient is negative means that darkness _lessens_ the 
likelihood that a stopped driver will be black, after adjusting for clock 
time. This matches our intuitive result from 6:30pm.

The fact that the standard error is small means that this is a _statistically
significant_ finding. Another important thing to do, besides checking 
statistical significance of your results, is to do a bunch of different 
robustness checks. The idea is that all models make some choices about how to 
turn the question at hand into math, and we want to vary those choices to make
sure that none of them change the model result -- that the result holds regardless
of, for example, how you choose to control for time. In our model we used a
natural spline with six degrees of freedom, but we could have chosen to use
3 degrees of freedom, or to do something way simpler like bin time to one-minute 
or five-minute windows.

Another robustness check that often comes up with this type of modeling is
adding in a control for sub-geography. For the outcome test, we controlled
for police district under the proposition that maybe it was an intentional
policy decision for certain police districts to lower search thresholds 
than other police districts -- and if those districts happened to be 
disproportionately black, that would "explain away" the bias we saw in the 
aggregate outcome test results.

In the case of the Veil of Darkness test, one could make an analogous claim:
Is the result that we're seeing just a product of certain districts being policed
more after dark and those districts have a lower proportion of black drivers 
than the districts being policed before dark? If so, then adjusting for location
in our model would lead to the darkness coefficient being zero. But if the
coefficient on darkness doesn't change much (i.e., stays negative in a 
statistically significant way), then that's another marker that our finding is
robust.

```{r}
mod2 <- glm(
  is_black ~ is_dark + splines::ns(minute, df = 6) + as.factor(district),
  family = binomial,
  data = vod_stops
)

summary(mod2)$coefficients["is_darkTRUE", c("Estimate", "Std. Error")]
# exp(summary(mod2)$coefficients["is_darkTRUE", "Estimate"])
```

Looks like no matter how you slice it, the veil of darkness test shows racial
profiling of black drivers is present in Philly traffic stops.
